Inverse energy transfer in three-dimensional quantum vortex flows

### 2025-11-17 main_REV.tex
Updated manuscript with revised figures

Remarkably, the same cascade phenomenology is observed in turbulent flows 
of quantum fluids, {\textit{i.e.} fluids at very low temperatures whose physics is
dominated by quantum effects.
Examples of such fluids are superfluid helium 
and atomic Bose-Einstein Condensates (BECs). 
The dynamics of these systems can be successfully depicted in terms of 
a two-fluid model \cite{tisza-1938,landau-1949,skrbek-sreenivasan-2012} 
describing the quantum fluid as the mixture of two components, 
the superfluid component and the thermal (or normal) component, which 
interact by means of a mutual friction force 
\cite{jackson-etal-2009,hall-vinen-1956a,hall-vinen-1956b}. 
The superfluid component flows without viscosity and
vanishing entropy; its vorticity is
confined to effectively one-dimensional vortex filaments
of atomic core thickness (called quantum vortices or vortex lines), 
around which the circulation of the velocity is quantised.
In BECs the thermal component forms a ballistic gas,
whereas in superfluid $^4$He it can be described as
a classical viscous fluid.
Despite these significant differences with respect to ordinary fluids, 
the direct
%forward 
kinetic energy cascade has
indeed been observed in three-dimensional superfluid
turbulence 
\cite{maurer1998,salort2010turbulent,baggaley2012,sherwin-robson2015,Muller_KolmogorovKelvinWave_2020,Muller_IntermittencyVelocityCirculation_2021}.
Evidence of this 
%forward 
direct
cascade has been found also in
three-dimensional turbulent BECs \cite{middleton-spencer2022}. 
In this forward energy cascade a fundamental role is played by the dynamics and interactions of quantum vortices,
in particular by their reconnections, phenomena where  
two vortex lines collide and recombine, exchanging heads and tails, 
altering the overall topology of the flow
\cite{koplik-levine-1993,bewley-etal-2008,rorai-etal-2016,serafini-etal-2017,galantucci-baggaley-parker-barenghi-2019,villoisUniversalNonuniversalAspects2017,villois2020irreversible}.
Reconnections occur frequently in 3D quantum turbulence and are essential to
the energy transfer towards small scales by engendering the the breakdown of coherent vortex structures
and triggering the Kelvin-wave cascade \cite{vinen-2001}.
In 2D, similarly to classical turbulence, 
an inverse energy cascade characterises turbulence in two-dimensional BECs, 
as shown in theoretical \cite{bradley2012energy,reeves2013,simula2014emergence,Muller_ExploringEquivalenceTwoDimensional_2024} and experimental \cite{johnstone2019evolution,gauthier2019giant} studies.

@@

In this work, we unveil a similar dynamics occurring in superfluid helium
($^4$He) as a result of vortex reconnections. 
%\red{The onset and development of quantum turbulence is fundamentally driven by such reconnections, which facilitates energy transfer and the breakdown of coherent vortex structures.}  
%Reconnections occur continuously in turbulence: they take place when
%two vortex lines collide and recombine, exchanging heads and tails, 
%altering the overall topology of the flow
%\cite{koplik-levine-1993,bewley-etal-2008,rorai-etal-2016,serafini-etal-2017,galantucci-baggaley-parker-barenghi-2019,villoisUniversalNonuniversalAspects2017,villois2020irreversible}. 
We show that the mutual friction force arising from the  
vortex reconnection is chiral, injecting in the normal fluid prevalently 
helicity of a given sign 
%\red{which is not globally conserved across the reconnection event}. 
Thus, as a consequence of vortex reconnections,
we observe an increase of the chiral imbalance of the quantum fluid, producing a transfer of kinetic energy from small to large scales, similarly to the phenomenology observed in 3D helically-decimated classical flows. 
 Unlike classical fluids, such a chiral imbalance arises naturally as physical process in the normal fluid.% because of vortex reconnections. 
%This is contrast to classical fluid dynamics,  where the helical characteristics of the flows producing 
% an inverse energy transfer require a careful synthetic construction 
% \cite{biferaleInverseEnergyCascade2012a,biferale-etal-2013,sahoo-etal-2017,plunianInverseCascadeEnergy2020a}.

@@

    %  \includegraphics[width=0.49\textwidth]{helicity_integralScale.pdf}
       \begin{subfigure}{0.49\textwidth}
         \centering
         \includegraphics*[width=\textwidth]{helicity_total.pdf}
    %     \caption{}\label{fig:}
     \end{subfigure}
     \begin{subfigure}{0.49\textwidth}
         \centering
         \includegraphics*[width=\textwidth]{energy-spec-scaled.pdf}
    %     \caption{}\label{fig:}

     \end{subfigure}
\caption{\CHANGE{ \emph(a): Temporal evolution of the total normal fluid helicity $\mathcal{H}$, normalised by the quantum of circulation $\kappa$. 
Bold solid lines refer to orthogonal reconnections, slightly faded lines refer to the set of Hopf-links reconnections. Red color refers to $T=1.9K$, blue to $T=2.1K$.
\emph(b): Normal fluid kinetic energy spectrum $E(k)$ before 
reconnection (dashed lines), at reconnection (solid lines) 
and after reconnection (dotted lines), for a pair of initially orthogonal vortices. Red (blue) lins correspond to  $T=1.9K$ ($T=2.1K$).}
%For superfluid helium, it is proper to make the helicity dimensionless in terms of $\kappa^2$.
}
%\red{The 3D rendering shows the asynchronous nature of the reconnections in the initial configuration, with the two reconnections occurring at different times; same colour scheme as in Fig.~\ref{fig:visualisation}.}                    }
\label{fig:total-helicity}
\end{figure}

@@

\begin{figure}
    \centering
    %\includegraphics[width=0.49\textwidth]{energySpec-fluxSpec.pdf}

     \begin{subfigure}{0.49\textwidth}
         \centering
         \includegraphics*[width=\textwidth]{flux-spec-scaled.pdf}
    %     \caption{}\label{fig:kinetic-energy}
     \end{subfigure}
     \begin{subfigure}{0.49\textwidth}
         \centering
         \includegraphics*[width=\textwidth]{integral-scale-scaled.pdf}
    %     \caption{}\label{fig:energy-flux}
     \end{subfigure}
%\begin{subcaptiongroup*}
%	\phantomsubcaption\label{fig:energy-spec}
%	\phantomsubcaption\label{fig:flux-spec}
%\end{subcaptiongroup*}

>>>>

Remarkably, the same cascade phenomenology is observed in turbulent flows 
of quantum fluids, {\textit{i.e.} fluids at very low temperatures whose physics is
dominated by quantum effects.
Examples of such fluids are superfluid helium 
and atomic Bose-Einstein Condensates (BECs). 
The dynamics of these systems can be successfully depicted in terms of 
a two-fluid model \cite{tisza-1938,landau-1949,skrbek-sreenivasan-2012} 
describing the quantum fluid as the mixture of two components, 
the superfluid component and the thermal (or normal) component, which 
interact by means of a mutual friction force 
\cite{jackson-etal-2009,hall-vinen-1956a,hall-vinen-1956b}. 
The superfluid component flows without viscosity and
vanishing entropy; its vorticity is
confined to effectively one-dimensional vortex filaments
of atomic core thickness (called quantum vortices or vortex lines), 
around which the circulation of the velocity is quantised.
In BECs the thermal component forms a ballistic gas,
whereas in superfluid $^4$He it can be described as
a classical viscous fluid.
Despite these significant differences with respect to ordinary fluids, 
the direct
%forward 
kinetic energy cascade has
indeed been observed in three-dimensional superfluid
turbulence 
\cite{maurer1998,salort2010turbulent,baggaley2012,sherwin-robson2015,Muller_KolmogorovKelvinWave_2020,Muller_IntermittencyVelocityCirculation_2021}.
Evidence of this 
%forward 
direct
cascade has been found also in
three-dimensional turbulent BECs \cite{middleton-spencer2022}. 
In this forward energy cascade a fundamental role is played by the dynamics and interactions of quantum vortices,
in particular by their reconnections, phenomena where  
two vortex lines collide and recombine, exchanging heads and tails, 
altering the overall topology of the flow
\cite{koplik-levine-1993,bewley-etal-2008,rorai-etal-2016,serafini-etal-2017,galantucci-baggaley-parker-barenghi-2019,villoisUniversalNonuniversalAspects2017,villois2020irreversible}.
Reconnections occur frequently in 3D quantum turbulence and are essential to
the energy transfer towards small scales by engendering the breakdown of coherent vortex structures
and triggering the Kelvin-wave cascade \cite{vinen-2001}.
In 2D, similarly to classical turbulence, 
an inverse energy cascade characterises turbulence in two-dimensional BECs, 
as shown in theoretical \cite{bradley2012energy,reeves2013,simula2014emergence,Muller_ExploringEquivalenceTwoDimensional_2024} and experimental \cite{johnstone2019evolution,gauthier2019giant} studies.

@@

In this work, we unveil a similar dynamics occurring in superfluid helium
($^4$He) as a result of vortex reconnections. 
%\red{The onset and development of quantum turbulence is fundamentally driven by such reconnections, which facilitates energy transfer and the breakdown of coherent vortex structures.}  
%Reconnections occur continuously in turbulence: they take place when
%two vortex lines collide and recombine, exchanging heads and tails, 
%altering the overall topology of the flow
%\cite{koplik-levine-1993,bewley-etal-2008,rorai-etal-2016,serafini-etal-2017,galantucci-baggaley-parker-barenghi-2019,villoisUniversalNonuniversalAspects2017,villois2020irreversible}. 
We show that the mutual friction force arising from the  
vortex reconnection is chiral, injecting in the normal fluid prevalently 
helicity of a given sign 
%\red{which is not globally conserved across the reconnection event}. 
Thus, as a consequence of vortex reconnections,
we observe an increase of the chiral imbalance of the quantum fluid, producing a transfer of kinetic energy from small to large scales, similarly to the phenomenology observed in 3D helically-decimated classical flows. 
 Unlike classical fluids, such a chiral imbalance arises naturally as a physical process in the normal fluid.% because of vortex reconnections. 
%This is contrast to classical fluid dynamics,  where the helical characteristics of the flows producing 
% an inverse energy transfer require a careful synthetic construction 
% \cite{biferaleInverseEnergyCascade2012a,biferale-etal-2013,sahoo-etal-2017,plunianInverseCascadeEnergy2020a}.

@@

    \includegraphics[width=0.49\textwidth]{helicity_and_energy-spectrum.pdf}
    %    \begin{subfigure}{0.49\textwidth}
    %      \centering
    %      \includegraphics*[width=\textwidth]{helicity_total.pdf}
    % %     \caption{}\label{fig:}
    %  \end{subfigure}
    %  \begin{subfigure}{0.49\textwidth}
    %      \centering
    %      \includegraphics*[width=\textwidth]{energy-spec-scaled.pdf}
    % %     \caption{}\label{fig:}

    %  \end{subfigure}
\caption{\CHANGE{ \emph(a): Temporal evolution of the total normal fluid helicity $\mathcal{H}$, normalised by the quantum of circulation $\kappa$. 
Bold solid lines refer to orthogonal reconnections, slightly faded lines refer to the set of Hopf-links reconnections. Red color refers to $T=1.9K$, blue to $T=2.1K$.
\emph(b): Normal fluid kinetic energy spectrum $E(k)$ before 
reconnection (dashed lines), at reconnection (solid lines) 
and after reconnection (dotted lines), for a pair of initially orthogonal vortices. Red (blue) lins correspond to  $T=1.9K$ ($T=2.1K$).}
%For superfluid helium, it is proper to make the helicity dimensionless in terms of $\kappa^2$.
}
%\red{The 3D rendering shows the asynchronous nature of the reconnections in the initial configuration, with the two reconnections occurring at different times; same colour scheme as in Fig.~\ref{fig:visualisation}.}                    }
\label{fig:total-helicity}
\end{figure}

@@

\begin{figure}
    \centering
    \includegraphics[width=0.49\textwidth]{flux_and_integral-scale.pdf}

    %  \begin{subfigure}{0.49\textwidth}
    %      \centering
    %      \includegraphics*[width=\textwidth]{flux-spec-scaled.pdf}
    % %     \caption{}\label{fig:kinetic-energy}
    %  \end{subfigure}
    %  \begin{subfigure}{0.49\textwidth}
    %      \centering
    %      \includegraphics*[width=\textwidth]{integral-scale-scaled.pdf}
    % %     \caption{}\label{fig:energy-flux}
    %  \end{subfigure}
%\begin{subcaptiongroup*}
%	\phantomsubcaption\label{fig:energy-spec}
%	\phantomsubcaption\label{fig:flux-spec}
%\end{subcaptiongroup*}
